---
title: "Exterior Temperature Calculations"
output: html_notebook
bibliography: ExtTempCalc.bib # build this collection in Zotero and save in this directory
---

This notebook is documentation of a method to model the heat transfer from the surfaces of the hot water cylinder to the ambient when tested under AS/NZS 4692.1.
  
The heat transfer is conductive from the water through the insulation to the surface per Appendix E of AS/NZS 4692.1.[@jointtechnicalcommitteeel-20electricwaterheatingappliancesElectricWaterHeaters2005a] For each surface, top, side, and bottom, assume the surface temperature is constant. Assume the water in the tank is perfectly mixed and at a constant temperature. Neglect the thermal resistance of the metal wall of the tank and at the surface of the insulation.

Calculate the heat transfer from the top and side surfaces with a combination of radiative and convective heat transfer to ambient. For the side the convective heat transfer is calculated with a combination of natural and forced convection. The heat transfer from the bottom is calculated as conductive through a wood platform and then convective downward through air and radiative to the floor. Assume all surfaces in the test chamber are at the ambient temperature.
  
The heat transfer from the water to the surfaces must equal the heat transfer from the surfaces to the test chamber. This is done by setting the initial surface temperatures to 25 °C then calculating the initial heat transfer to and from the surfaces. The surface temperatures are then iteratively adjusted until the heat fluxes at each surface are equal.

```{r}
Tamb   = 20  # °C
Twater = 75  # °C
Vair   = 0.25  # air speed within range of 0.25 m/s and 0.5 m/s
k_ins  = 0.025  # polyurethane W/m-K, from 4234 Appendix E
```

The water heater and information about it are from the Predictive Modeling Tool (PMT) @bartmilneOptimisationStatisticalEstimation2021 The insulation thicknesses determined by the PMT are used here.

This example is based on the Rinnai model 'MS1805503 0'  tested to 4692. For volumes, dimensions, and insulation thicknesses see Table 7: Hot water cylinder model, MEPS and measurement information. 

```{r}
vol       =	0.18  # m^3	
diam      =	0.55  # m
height		= 1.295 # m
dx_side		= 0.053 # m side insulation thickness
dx_ends		= 0.060 # m top and bottom insulation thickness

```

These calculations are based on the heat transfer equations and correlations from from Chapter 4, Heat Transfer, ASHRAE Fundamentals.@technicalcommittee1.3heattransferandfluidflow.ChapterHeatTransfer2017 

```{r}
source("ExtTempFunc.R")
```

## Conductive heat loss from water to surface #

The simple idealized tank model in 4234 Appendix E consists only of losses through the insulation on the sides and at the ends of the tank. There are no losses through the corners. Thus the insulation can be modeled as a hollow cylinder and two circular disks.

To check the equations at this step in the calculations assume the initial surface temperature of the insulation at each side is 25 °C 

```{r}
Tsurf.top  = 25 # °C 
Tsurf.side = 25 # °C 
Tsurf.bot  = 25 # °C 
```


### Conductive heat loss through side insulation #

For the hollow cylinder the equation is written as,
$$Q_{surf.side} = \frac{2 \pi k_{ins} h}{ln(\frac{d_o}{d_i})} {\Delta T } $$
Or rewritten to account for the thicknesses of insulation,
$$Q_{surf.side} = \frac{2 \pi k (height- 2 dx_{ends} )}
{ln(\frac{diam}{(diam-(2dx_{side}))})}{(T_{water} - T_{surf.side})) } $$



```{r}
# calculate Qsurf.side
Qsurf.side <- Qcond.cyl(length = height - 2 * dx_ends,
                       r_inner = diam-(2*dx_side),
                       r_outer = diam,
                       T_inner = Twater,
                       T_outer = Tsurf.side,
                       k = k_ins)
```


So with the initial surface temperature, the heat loss through the side insulation is `r round(Qsurf.side,1)` (W) or in kWh/day `r round(Qsurf.side*(24/1000),3)`

### Conductive heat loss through top and bottom insulation #

The conductive loss through the top is

$$
Q_{surf.top} = \frac{ k_{ins} \pi  (r_i)^2 } 
{dx_{ends}}
 \Delta T
$$
Which can be rewritten as 

$$
Q_{surf.top} = \frac{ \pi k_{ins}  \left( \frac{diam - 2 dx_{side}}{2} \right)^2  }
{dx_{ends}}(T_{water} - T_{surf.top})
$$

```{r}
Qsurf.top = Qcond.disc(radius = ((diam - 2 * dx_side)/2), 
                       thick = dx_ends,
                       Tcold = Tsurf.top)

```

So with the initial surface temperature, the conductive heat loss through the insulation at the ends is `r round(Qsurf.top,1)` (W) or `r round(Qsurf.top*(24/1000),3)` (kWh/day)


Similarly the heat transfer through the insulation to the bottom surface is

$$
Q_{surf.bot} = \frac{ \pi k_{ins}  \left( \frac{diam - 2 dx_{side}}{2} \right)^2  }
{dx_{ends}}(T_{water} - T_{surf.bot})
$$
```{r}
Qsurf.bot = Qcond.disc(radius = ((diam - 2 * dx_side)/2), 
                       thick = dx_ends,
                       Tcold = Tsurf.bot)

```

With the same initial temperatures this is also `r round(Qsurf.bot,1)` (W) or `r round(Qsurf.bot*(24/1000),3)` (kWh/day)

# Heat Loss from Surface to Ambient #

The heat losses from the surface of the water heater will a combination of convective, radiative and for the base further conduction. 

Losses from the side surfaces are modeled as convective to the ambient air and radiative to the surfaces of the test chamber. The convective losses will be modeled as a combination of free and forced convection from a heated vertical surface...

Losses from the top are modeled as free convection upward from a heated disk and also radiative.

Loss from the bottom is modeled as conductive through a wooden platform and then radiative to the test chamber surfaces and downward convection through the air below the platform.


Convective heat transfer is often calculated by determining a Nusselt number (Nu) for the conditions being investigated. Nu is the ratio of convective to conductive heat transfer at a boundary with a fluid. The conductive component is measured under the same conditions but for a hypothetically motionless fluid. So Nu = hLc/k, where Lc is a characteristic length. Solving for h, the convective heat transfer coefficient, then allows the heat loss to be calculated. 

Nusselt numbers are usually obtained from correlations of experimental data. The correlations to determined Nu are written in terms of Prandtl numbers and Rayleigh numbers. In the case of force convection the Reynolds number is used instead of the Rayleigh number.

The Rayleigh number (Ra) for a fluid is a dimensionless number associated with buoyancy-driven flow, also known as free or natural convection. It characterises the fluid's flow regime: a value in a lower range denotes laminar flow; a value in a higher range, turbulent flow. Below a critical value, there is no fluid motion and heat transfer is by conduction rather than convection. The Rayleigh number is defined as the product of the Grashof number and the Prandtl number.

The Grashof number (Gr) is a dimensionless number in fluid dynamics and heat transfer which approximates the ratio of the buoyancy to viscous force acting on a fluid. It frequently arises in the study of situations involving natural convection and is analogous to the Reynolds number.

The Prandtl number (Pr) is a dimensionless number defined as the ratio of momentum diffusivity to thermal diffusivity.

#### Grashof Number #

The Grashof number is calculated as 

$$Gr_ = \frac{g \beta (T_s-T_{\infty}) L^3}{\mu^2}$$

where;
```{r}
g     = 9.81     # acceleration due to Earth's gravity (m/s^2)
beta  = 3.43e-3  # the coefficient of thermal expansion of air at 20°C (1/K)
Ts    = 25       # the surface temperature (°C ), using initial estimate here
Tamb  = 20       # the bulk temperature (°C)
mu    = 15.28e-6 # the kinematic viscosity (m2/s), using air at 22.5°C

```

```{r}
Gr <- Gr(L=height, Tsurf = Ts)

```

With these values the Grashof number is `r sprintf("%g", Gr)`

If the diameter of vertical cylinder is large enough relative to the height, the natural convection correlations for a vertical heated plate can be used.@technical_committee_13_heat_transfer_and_fluid_flow_chapter_2017 Using this correlation is appropriate when $$ D/L > 35/Gr^{1/4} $$

In this case $\frac{D}{L}$ is `r sprintf("%5.4f",diam/height)` and $\frac{35}{Gr^{1/4}}$ is `r sprintf("%5.4f",35/(Gr^(1/4)))` so treating a hot water cylinder as a vertical plate for natural convection is appropriate.

#### Prandtl number #

The Prandtl number is calculated as 

$$Gr = \frac{c_p \mu}{k}= \frac{\nu}{\alpha}$$

where;
```{r}
mu    = 15.28e-6  # the kinematic viscosity (m2/s), for air air at 22.5°C
alpha = 22.045e-6 # the thermal diffusivity (m2/s) for air at 22.5°C)

```


```{r}
# Calculate the current value of Pr
Pr <- Pr(mu, alpha)
```
With these values the Prandtl number is `r sprintf("%7.4f ", Pr)`


#### Rayleigh number #

The Rayleigh number is calculated as ${Gr}{Pr}$. 


```{r}
# Calculate the current value of Ra
Ra <- Ra(Gr,Pr)
```

In these conditions the Rayleigh number is `r sprintf("%5.4e",Ra)`


## Heat Loss from Tank Side Surface to Ambient #

### Natural Convection #

For this Rayleigh number, the Nusselt Number for natural convection from a vertical plate is calculated as @technical_committee_13_heat_transfer_and_fluid_flow_chapter_2017 
$$
Nu_{vert} = \left\{{0.825 + \frac{0.387Ra^{\frac{1}{6}}}
                { \left[ 1 + \left( \frac{0.437}{Pr} \right) ^ \frac{9}{16} 
                \right] ^\frac{8}{27}}} 
                \right\}^2
$$


```{r}
Nu.vert.nat <- Nu.vert.nat(Ra,Pr)
```

For these conditions the Nusselt number of the natural convective off the side of the tank is `r sprintf("%5.2f",Nu.vert.nat)`

To apply the Nusselt number requires the thermal conductivity of air

```{r}
k_air  = 26.06e-3 #  W/m-K thermal conductivity of air at 22.5 °C      
```

The heat transfer coefficient is the $ h = Nu * k / height$

```{r}
# calculate the heat transfer coefficient in W/m^2-K
h_side.conv.nat <- Nu.vert.nat * k_air / height

```

So the convective heat transfer coefficient from the side of the tank is `r sprintf("%5.2f",h_side.conv.nat)` W/m^2-k 

The heat transfer from natural convection off the side of the tank is
$$
Q_{side.conv} = h_{side} ( diam  \times \pi \times height)  (T_{surf}-T_{amb})
$$
```{r}
 Qside.conv.nat <- h_side.conv.nat * diam * pi * height * (Tsurf.side - Tamb)
```

So the natural convective heat loss from the side of the tank is
`r sprintf("%5.3f",Qside.conv.nat)` W or `r sprintf("%5.3f",Qside.conv.nat*24/1000)` kWh/day
`

### Forced Convection #

The Nusselt Number for forced convection across a cylinder is calculated as @technical_committee_13_heat_transfer_and_fluid_flow_chapter_2017 
$$
Nu_{cyl} =  0.3 + \frac{0.62 Re^{\frac{1}{2}}Pr^{\frac{1}{3}}}
                  { \left[ 1 + (\frac{0.4}{Pr})^{\frac{2}{3}} \right] ^{\frac{1}{4}}}
\left[ 1 + \left( \frac{Re}{282000} \right) ^ {\frac{5}{8}} \right] ^{\frac{4}{5}}
$$

The Reynolds number (Re) for cross flow over cylinder is $ \frac{VD}{\nu}$ 
where;
  V = air speed velocity, m/s  
  D = diameter, m   (This is the charactistic number here)
  $\nu$ = kinematic viscosity of air, m^2/s


```{r}
# Calculate the current value of Re
Re.cyl.frc <-Re.cyl.frc(
  V = Vair,
  D = diam,
  nu = nu()
)
```



```{r}
Nu.vert.nat <- Nu.vert.nat(Ra,Pr)
```

For these conditions the Nusselt number of the natural convective off the side of the tank is `r sprintf("%5.2f",Nu.vert.nat)`

To apply the Nusselt number requires the thermal conductivity of air

```{r}
k_air  = 26.06e-3 #  W/m-K thermal conductivity of air at 22.5 °C      
```

The heat transfer coefficient is the $ h = Nu * k / height$

```{r}
# calculate the heat transfer coefficient in W/m^2-K
h_side.conv.nat <- Nu.vert.nat * k_air / height

```

So the convective heat transfer coefficient from the side of the tank is `r sprintf("%5.2f",h_side.conv.nat)` W/m^2-k 

The heat transfer from natural convection off the side of the tank is
$$
Q_{side.conv} = h_{side} ( diam  \times \pi \times height)  (T_{surf}-T_{amb})
$$
```{r}
 Qside.conv.nat <- h_side.conv.nat * diam * pi * height * (Tsurf.side - Tamb)
```

So the natural convective heat loss from the side of the tank is
`r sprintf("%5.3f",Qside.conv.nat)` W or `r sprintf("%5.3f",Qside.conv.nat*24/1000)` kWh/day
`






```{r}
Qsurf.side
Qside.conv.nat
```
Qsurf.side > Qside.conv so have to raise Tsurf until they are equal.
except probably need to add Qside.conv.frc



# References
